Davos: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2023-09-25 paper/main.tex
WIP updating illustrative example text

@@

The Davos package is named after Davos Seaworth, a smuggler referred
to as ``the Onion Knight'' from the series \textit{A Song of Ice and Fire} by
George R. R. Martin~\cite{Mart98}. The \texttt{smuggle} keyword provided by
Davos is a play on Python's \texttt{import} keyword: whereas importing
can load a package into the Python workspace within the existing rules and
frameworks provided by the Python language, ``smuggling'' provides an
alternative that expands the scope and reach of ``importing.'' Like the
character Davos Seaworth (who became famous for smuggling onions through a
blockade on his homeland), we use ``onion'' comments to precisely control how
packages are smuggled into the Python workspace.

@@

\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Next, in lines 8--9 we \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 4 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Standard library modules are included with all Python
distributions, so this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 5 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring that users have a particular exact version
installed would likely be unnecessarily restrictive. However, a
\textit{future} release might introduce some breaking change.  The
onion comment in line 5 helps ensure the analysis notebook continues
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
Davos's \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}
Setting the \texttt{auto\_rerun} attribute to \texttt{True} is particularly useful
for managing the installation of \texttt{pandas} in the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example4}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a pickled \texttt{Panel} object, then we must use a version of
\texttt{pandas} prior to 0.25.0 (i.e., the version in which the \texttt{Panel}
class was removed) to be able to load it in. Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, written to a CSV
file, which is compatible with newer \texttt{pandas} versions).

@@

>>>>

\newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO}: #1}}
\newcommand{\stoppedhere}{\bigskip\bigskip\textcolor{red}{\textbf{========== TODO: finish editing from here to end ==========}}\bigskip\bigskip}

@@

The Davos package is named after Davos Seaworth, a smuggler referred
to as ``the Onion Knight'' from the series \textit{A Song of Ice and Fire} by
George R. R. Martin~\cite{Mart98}. The \texttt{smuggle} keyword provided by
Davos is a play on Python's \texttt{import} keyword: whereas importing
can load a package into the Python workspace within the existing rules and
frameworks provided by the Python language, ``smuggling'' provides an
alternative that expands the scope and reach of ``importing.'' Like the
character Davos Seaworth (who became famous for smuggling onions through a
blockade on his homeland), the Davos package uses ``onion comments'' to precisely control how
packages are smuggled into the Python workspace.

@@

After installing and importing Davos (lines 1--2), we first use the \texttt{davos.require\_python()} function to constrain the Python version used to run the code below (see Sec.~\ref{subsec:toplevel}).
As described above, the example code in Figure \ref{fig:illustrative-example} uses two different versions of the \texttt{pandas} library: first, an older one needed to access a dataset saved in an outmoded format, then a newer one throughout the remainder of the notebook.
Line 6 therefore ensures upfront that the notebook's Python version falls within the overlap between the ranges of Python versions that these two versions of \texttt{pandas} support. If it does not, the function in line 6 will raise an error that includes a message to this effect (lines 4--5).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example1}
\end{center}
Next, in lines 8--9 we \texttt{smuggle} two
utilities for interacting with local files in the code below. The
\texttt{smuggle} statement in line 8 loads the \texttt{is\_file()}
function from the Python standard library's \texttt{os.path}
module. Since standard library modules are included with all Python
distributions, this line is functionally equivalent to an
\texttt{import} statement and does not need or benefit from an onion
comment. Line 9 loads the \texttt{joblib} package~\cite{Varo10},
installing it first, if necessary. Since \texttt{joblib}'s I/O
interface has historically remained stable and backwards-compatible
across releases, requiring a particular exact version
would likely be unnecessarily restrictive. However, it is possible a
\textit{future} release could introduce some breaking change.  The
onion comment in line 9 helps ensure the analysis notebook continues
to run properly in the future by limiting allowable versions to those
already released when the code was written:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example2}
\end{center}
Line 7 then uses the \texttt{davos.config} object to enable
Davos's \texttt{auto\_rerun} option before smuggling the next
two packages: \texttt{NumPy} and
\texttt{pandas}. Because these packages rely heavily on custom C data
types, loading the particular versions from the onion comments may
require restarting the notebook kernel if different versions had been previously
imported during the same interpreter session (see
Sec.~\ref{subsec:config}).
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example3}
\end{center}

\stoppedhere

Setting the \texttt{auto\_rerun} attribute to \texttt{True} is particularly useful
for managing the installation of \texttt{pandas} in the next
lines:
\begin{center}
\includegraphics[width=0.9\textwidth]{figs/example4}
\end{center}
If we suppose that the data contained in \texttt{data-old.pkl} is
stored in a pickled \texttt{Panel} object, then we must use a version of
\texttt{pandas} prior to 0.25.0 (i.e., the version in which the \texttt{Panel}
class was removed) to be able to load it in. Line 11 ensures
that an older version of \texttt{pandas} will be imported, enabling
the data to be read in (and, in line 13, written to a CSV
file, which is compatible with newer \texttt{pandas} versions).

@@

\todo{mention notebook reproducibility, cell order, multiple package versions re: reviewer's comment; note importance of running lines 14--19 \& 38--40 in single cell; say another option is to alias one version, e.g., \texttt{smuggle pandas as pd\_old}}
